---
title: "Simulate time saved nationally by a simplified ethics/SSA system"
author: "Adrian Barnett"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: word_document
---

```{r setup, include=FALSE}
# basic set up
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE, comment='', dpi=400)
options(width=1000, scipen = 999) # Wide pages and no scientific numbers

# libraries
library(ggplot2)
library(dplyr)
library(tidyverse)

## key data
# from NHMRC report
n.proposals = 18039 # number of new proposals per year (not re-submissions or amendments); from 195 committees
# high risk
n.trials = 2155 # new clinical trial
n.gametes = 0 # hard to know
n.atsi = 0 # hard to know
p.low.neg = 0.4 # proportion of all applications that are low and negligible risk
p.need.SSA = 0.4 # proportion of low and negligible risk applications that need an SSA
# multiply together to get the number we consider
n.lowrisk.applications = round(n.proposals*p.low.neg*p.low.neg)

# get empirical data on times
source('empirical.data.R')

```

We model the national number of low and negligible risk ethics applications and estimate: 1) the time spent currently and, 2) how much time could be saved in a streamlined system.

# Time spent in the current system

We start with the total number of ethics proposals from an NHMRC report. 
This was `r format(n.proposals, big.mark=',')` in 2016.

The proportion of low and negligible risk applications was not given in the report, hence we estimate it to be `r p.low.neg` (_IMPROVE_).

We also want to consider applications that need an SSA, so those would include healthcare facilities.
We estimate that this proportion would be `r p.need.SSA` (_IMPROVE_), and this is the proportion of all low and negligible risk applications.

This means the overall proportion of low and negligible risk applications is `r p.low.neg*p.low.neg` and the total number of applications we consider is `r format(n.lowrisk.applications, big.mark=',')` per year.

```{r simulate, include=FALSE}
## number of committees ##
# split into ethics ...
ethics.frame = data.frame(id = 1:n.lowrisk.applications, n.ethics= 1 + rpois(lambda=0.5, n=n.lowrisk.applications)) %>% # number of ethics commitees, minimum of 1
  uncount(n.ethics) %>% # expand rows for each committee
  mutate(time = sample(ethics.times, replace = TRUE, size=n())) %>%
  group_by(id) %>%
  summarise(n.ethics = n(), total.ethics = sum(time), time.ethics=max(time)) %>% # total time and longest ethics committee
  ungroup()
# ...  and site-specific
ssa.frame = data.frame(id = 1:n.lowrisk.applications, n.ssa= 1 + rpois(lambda=2, n=n.lowrisk.applications)) %>% # number of ethics commitees, minimum of 1
  uncount(n.ssa) %>% # expand rows for each SSA
  mutate(time = sample(ssa.times, replace = TRUE, size=n())) %>%
  group_by(id) %>%
  summarise(n.ssa = n(), total.ssa = sum(time), time.ssa=max(time)) %>% # total time and longest SSA
  ungroup()
# now merge and calculate total hours per study
totals = full_join(ethics.frame, ssa.frame, by='id') %>%
  mutate(total.time = total.ethics + total.ssa, # total number of hours
         time = time.ethics + time.ssa) # time for approval (slowest time) which is the time for researchers
```


# Simulated number of ethics committees per proposal

We simulate the number of ethics committees and site-specific approvals needed per study using Poisson distributions, with a minimum of one ethics and one SSA approval.

```{r ethics.plot}
to.plot = select(totals, id, n.ethics, n.ssa) %>%
gather(n.ethics, n.ssa, key='app', value='number') %>%
mutate(facet = ifelse(app=='n.ethics', 'Ethics', 'Site-specific'))

hplot = ggplot(to.plot, aes(x=number)) +
  geom_bar()+
  scale_x_continuous(breaks=1:12)+
  xlab('Number of approvals per study')+
  ylab('Frequency')+
  theme_bw()+
  facet_wrap(~facet)+
  theme(panel.grid.minor = element_blank())
hplot
```

# Time spent per approval

For the times per approval, we use data from Barnett et al (2016) and Duplancic et al (2019) from their low and negligible risk studies.
We simulate times from these studies using an empirical distribution.
The observed times for ethics and SSA are shown below.

```{r bar, fig.width=11, fig.height=6}
plot = ggplot(studies, aes(x=id, y=time, fill=factor(which)))+
  geom_bar(stat='identity')+
  scale_fill_manual('', values=c('dark red','dark blue'))+
  theme_bw()+
  xlab('')+
  ylab('Time, days')+
  facet_wrap(~study, scales='free_x')
plot
```


# Simulated times

The plot below shows the simulated times in days for studies to get ethics and SSA approval.
The dotted vertical line is at six months.
The approval times are calculated using the longest HREC and longest SSA per study.

```{r time.plot}
hplot = ggplot(totals, aes(x=time)) +
  geom_histogram(bins=25)+
  xlab('Time in days')+
  ylab('Frequency')+
  theme_bw()+
  geom_vline(xintercept = 183, lty=2) # line at six months
hplot
```

# Total time and potential time saved

In the simplified system we assume all applications take no longer than six months, which is based on the system proposed by Clay-Williams et al (2019).

```{r times, include=FALSE}
update = mutate(totals,
                time.saved = time - 183, # minus 183 days for six months
                time.saved = ifelse(time.saved < 0, 0, time.saved)) # ignore faster approvals for time saved
# time in current system
total.pre = sum(update$time)
total.pre.years = round(sum(update$time) / 365.25) # convert to years including weekends
# time saved for researchers
total.post = sum(update$time.saved)
total.post.years = round(sum(update$time.saved) / 365.25) # convert to years including weekends
saved = total.pre.years - total.post.years
# salary (from Duplancic)
salary = 56000 # AUD
```

The estimated total time spent in the current system is `r format(total.pre.years, big.mark=',')` years.
The estimated total time spent in the streamlined system is `r format(total.post.years, big.mark=',')` years.
The number of years that could be saved is `r format(saved, big.mark=',')` years.

If we apply a full-time researchers offers time at an annual salary of \$`r format(salary, big.mark=',')`, then we have a salary cost of \$`r format(round(saved*salary/(10^6)), big.mark=',')` million per year.

# References

* Duplancic, C. , Crough, T. and Bell, S. C (2019) Multi‐centre ethics and research governance review can impede non‐interventional clinical research. _Intern Med J_, __49__: 722-728. doi:10.1111/imj.14158

* Robyn Clay-Williams, Natalie Taylor and Jeffrey Braithwaite (2019) Potential solutions to improve the governance of multicentre health services research. _Med J Aust_ __208__ (4). doi:10.5694/mja16.01268 

* National Health and Medical Research Council. Report on the Activity of Human Research Ethics Committees and Certified Institutions for the period: 1 January 2016 to 31 December 2016

* Adrian G. Barnett, Megan J. Campbell, Carla Shield, Alison Farrington, Lisa Hall, Katie Page, Anne Gardner, Brett G. Mitchell & Nicholas Graves (2016) The high costs of getting ethical and site-specific approvals for multi-centre research. _Research Integrity and Peer Review_ __1__ 16
