---
title: "Simulate time saved nationally by a simplified ethics/site-specific approval system"
author: "Adrian Barnett"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: word_document
---

```{r setup, include=FALSE}
# basic set up
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE, error=FALSE, comment='', dpi=400)
options(width=1000, scipen = 999) # Wide pages and no scientific numbers

# libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
set.seed(7654321)

## key data
# from NHMRC report
n.proposals = 18039 # number of new proposals per year (not re-submissions or amendments); from 195 committees
# number of trials
n.trials.nhmrc = 2155 # new clinical trial
p.nhmrc = n.trials.nhmrc / n.proposals
# QUT
n.qut.numerator = 110+70 # committee plus admin review
n.qut.denominator = 672 # total for 2018 
p.qut = n.qut.numerator/n.qut.denominator
n.qut = round(p.qut * n.trials.nhmrc) # extrapolate QUT number nationally
# from search of ANZCTR
n.anzctr = 1472
# from pubmed search of all clinical trials with an Australian affiliation in 2018
# 2018[PDAT] AND Australia[Affiliation] AND (clinical trial[Publication Type] OR Clinical Study[Publication Type] OR Clinical Trial[Publication Type] OR Clinical Trial, Phase I[Publication Type] OR Clinical Trial, Phase II[Publication Type] OR Clinical Trial, Phase III[Publication Type] OR Clinical Trial, Phase IV[Publication Type] OR Clinical Trial Protocol[Publication Type] OR Adaptive Clinical Trial[Publication Type] OR Equivalence Trial[Publication Type] OR Multicenter Study[Publication Type] OR Pragmatic Clinical Trial[Publication Type] OR Randomized Controlled Trial[Publication Type])
n.pubmed = 2942

# average to get number of low risk
n.lowrisk.applications = round( (n.qut + n.trials.nhmrc + n.anzctr + n.pubmed)/4 )

# get empirical data on times
source('empirical.data.R')
```

# 1. Approvals requiring ethics and SSA

We model the national number of applications requiring ethics and site-specific approvals and estimate: 1) the time spent currently and, 2) how much time could be saved in a streamlined system.

## 1.1 Time spent in the current ethics/SSA system

We estimate number of applications requiring ethics and SSA from four sources:

1. The number of trials in the NHMRC report on HREC activity was `r n.trials.nhmrc`.

2. The number of committee and administrative reviews at QUT in 2018 was `r n.qut.numerator` out of `r n.qut.denominator` applications, which gives a number of `r n.qut` if extrapolated to the national number of applications reported in the NHMRC report.

3. Searching on ANZCTR for studies registered in 2018 that included sites in Australia gives `r n.anzctr`. 

4. Searching on pubmed for clinical trials (of all types) with an Affiliation of "Australia" published in 2018 gives `r n.pubmed` papers.

We average these four numbers to give `r format(n.lowrisk.applications, digits=2)` applications requiring ethics and SSA per year.

```{r simulate, include=FALSE}
## number of committees ##
# split into ethics ...
ethics.frame = data.frame(id = 1:n.lowrisk.applications, n.ethics= 1 + rpois(lambda=0.4, n=n.lowrisk.applications)) %>% # number of ethics commitees, minimum of 1
  uncount(n.ethics) %>% # expand rows for each committee
  mutate(time = sample(ethics.times, replace = TRUE, size=n())) %>%
  group_by(id) %>%
  summarise(n.ethics = n(), total.ethics = sum(time), time.ethics=max(time)) %>% # total time and longest ethics committee
  ungroup()
# ...  and site-specific
ssa.frame = data.frame(id = 1:n.lowrisk.applications, 
  # number of ethics commitees, minimum of 1, use two Poisson to capture long tail 80:20 split
  big.small = rbinom(n=n.lowrisk.applications, size=1, prob=0.8), # P trial is small; big or small number of sites
  small.size = rpois(lambda=1.5, n=n.lowrisk.applications),
  big.size = rpois(lambda=10, n=n.lowrisk.applications)) %>%
  mutate(n.ssa = round(1 + big.small * small.size + (1-big.small) * big.size )) %>% 
  uncount(n.ssa) %>% # expand rows for each SSA
  mutate(time = sample(ssa.times, replace = TRUE, size=n())) %>%
  group_by(id) %>%
  summarise(n.ssa = n(), total.ssa = sum(time), time.ssa=max(time)) %>% # total time and longest SSA
  ungroup()
# now merge and calculate total hours per study
totals = full_join(ethics.frame, ssa.frame, by='id') %>%
  mutate(total.time = total.ethics + total.ssa, # total number of hours
         time = time.ethics + time.ssa) # time for approval (slowest time) which is the time for researchers
```

## 1.2 Simulated number of ethics committees and sites per proposal

We simulate the number of ethics committees and site-specific approvals needed per study using Poisson distributions, with a minimum of one ethics and one SSA approval. The site-specific approval needs a long right tail to cover those studies that recruit across multiple states. We therefore used a mixture of Poisson distributions to capture this, assuming 80% of studies are small and 20% are large.

```{r ethics.plot}
to.plot = select(totals, id, n.ethics, n.ssa) %>%
gather(n.ethics, n.ssa, key='app', value='number') %>%
mutate(facet = ifelse(app=='n.ethics', 'Ethics', 'Site-specific'))
hplot = ggplot(to.plot, aes(x=number)) +
  geom_bar()+
  scale_x_continuous(breaks=c(1,5,10,15,20))+
  xlab('Number of approvals per study')+
  ylab('Frequency')+
  theme_bw()+
  facet_wrap(~facet)+
  theme(panel.grid.minor = element_blank())
hplot
```

## 1.3 Time spent per approval in the current ethics/SSA system

For the times per approval, we use data from Barnett et al (2016) and Duplancic et al (2019) from their studies that required an ethics and site-specific approval.
We simulate times from these studies using an empirical distribution.
The observed times for ethics and SSA are shown below, each bar is an individual HREC or site.

```{r bar, fig.width=11, fig.height=6}
plot = ggplot(studies, aes(x=id, y=time, fill=factor(which)))+
  geom_bar(stat='identity')+
  scale_fill_manual('', values=c('dark red','dark blue'))+
  theme_bw()+
  xlab('')+
  scale_x_continuous(labels=NULL)+
  ylab('Time, days')+
  facet_wrap(~study, scales='free_x')+
  theme(text = element_text(size=14))
plot
```

## 1.4 Simulated approval times in the current ethics/SSA system

The plot below shows the simulated times in days for studies to get ethics and SSA approval.
The dotted vertical line is at six months.
The approval times are calculated using the longest HREC and longest SSA per study.

```{r time.plot}
hplot = ggplot(totals, aes(x=time)) +
  geom_histogram(bins=25)+
  xlab('Time in days')+
  ylab('Frequency')+
  theme_bw()+
  geom_vline(xintercept = 183, lty=2) # line at six months
hplot
```

## 1.5 Time in the simplified ethics/SSA system

In the simplified system we assume all applications take no longer than six months, which is based on the system proposed by Clay-Williams et al (2019). Hence we expect the tail of the distribution to be at six months as per the plot below. This is the same distribution as above squeezed into a shorter span.

```{r time.plot.six.months}
divisor = max(totals$time) / 183 # take the longest time and make it 6 months
hplot = ggplot(totals/divisor, aes(x=time)) +
  geom_histogram(bins=25)+
  xlab('Time in days')+
  ylab('Frequency')+
  theme_bw()+
  geom_vline(xintercept = 183, lty=2) # line at six months
hplot
```

## 1.6 Total time and potential time saved in the ethics/SSA system

```{r times, include=FALSE}
update = mutate(totals,
      time.new = time / divisor,
      time.saved = time - time.new #
      # simpler version using everything over 6 months
    #  time.saved = time - 183, # minus 183 days for six months
    #  time.saved = ifelse(time.saved < 0, 0, time.saved) # ignore faster approvals for time saved
                ) 
# time in current system
total.pre = sum(update$time)
total.pre.years = round(sum(update$time) / 365.25) # convert to years including weekends
# time saved for researchers
total.post = sum(update$time.saved)
total.post.years = round(sum(update$time.saved) / 365.25) # convert to years including weekends
saved = total.pre.years - total.post.years
# salary (from Duplancic)
salary = 56000 # AUD
```

The estimated total time spent in the current system is `r format(total.pre.years, big.mark=',')` years.
The estimated total time spent in the streamlined system is `r format(total.post.years, big.mark=',')` years.
The number of years that could be saved is `r format(saved, big.mark=',')` years.

If we apply a full-time researchers assistant's time at an annual salary of \$`r format(salary, big.mark=',')`, then we have a salary cost of \$`r format(round(saved*salary/(10^6)), big.mark=',')` million per year.

# 2. Low and negligible risk proposals

```{r numbers, include=FALSE}
# extrapolating QUT data
n.negligible.low.risk = 492
p.qut.low = n.negligible.low.risk/n.qut.denominator
n.national.estimate1 = round(p.qut.low*n.proposals)

## number of low risk from pubmed
# searches
#2018[PDAT] AND Australia[Affiliation] AND (Observational Study[PT] OR Evaluation Studies[PT] OR Observational Study, Veterinary[PT] OR Validation Studies[PT])
#2018[PDAT] AND (QUT[Affiliation] OR Queensland University of Technology[Affiliation]) AND(Journal Article[PT]) # 1628
n.qut.articles = 1628
#2018[PDAT] AND Australia[Affiliation] AND Journal Article[PT] # 50577
n.total.articles.australia = 50577
qut.prop = n.qut.articles / n.total.articles.australia
#
n.national.estimate2 = round((1/qut.prop) * n.negligible.low.risk)

# average
n.national.low = round((n.national.estimate1 + n.national.estimate2)/2)
```

Here were examine the potential savings for using an US-style self-certification system in place of the current low and negligible risks system.

We estimate the national number of low and negligible risk proposals in two ways.

1. By multiplying the national number of HREC applications (`r format(n.proposals, big.mark=',')`) by the proportion of low and negligible risk applications from QUT's HREC (`r format(p.qut.low, digits=2)`). This gives a total national number of low and negligible risk applications of `r format(n.national.estimate1, big.mark=',')`.

2. Using Pubmed, the proportional of journal articles contributed by QUT in 2018 of the total number of articles with Australian affiliations is `r format(qut.prop, digits=2)`. Using the inverse of this proportion applied to the number of low risk ethics applications at QUT in 2018 gives an estimated national number of `r format(n.national.estimate2, big.mark=',')`.

Averaging the two estimates gives a national estimate of `r format(n.national.low, big.mark=',')`.

## 2.1 Simulating the current approval times for low and negligible risk proposals 

We simulate the current times spent by researchers on low and negligible risk proposals using the empirical data on ethics proposals from the Barnett and Duplancic studies.

```{r times.low}
times.low = data.frame(id=1:n.national.low, time=sample(ethics.times, size=n.national.low, replace=TRUE))
tlow = ggplot(times.low, aes(x=time)) +
  geom_histogram()+
  xlab('Time in days')+
  ylab('Frequency')+
  theme_bw()+
  theme(panel.grid.minor = element_blank())
tlow
```

The median time is `r median(times.low$time)` days, and the inter-quartile range is `r quantile(times.low$time, 0.25)` to `r quantile(times.low$time, 0.75)` days.

## 2.2 Time in the low and negligible risk proposals 

We assume that a self-certification system would take researchers 2 days per proposal.

```{r}
update.low = mutate(times.low,
      time.new = 2, # just two days in new system
      time.saved = time - time.new) 
# time in current system
total.pre.low = sum(update.low$time)
total.pre.low.years = round(sum(update.low$time) / 365.25) # convert to years including weekends
# time saved for researchers
total.post.low = sum(update.low$time.new)
total.post.low.years = round(sum(update.low$time.new) / 365.25) # convert to years including weekends
saved.low = total.pre.low.years - total.post.low.years # time saved in years
```

The estimated total time spent in the current system is `r format(total.pre.low.years, big.mark=',')` years.
The estimated total time spent in the streamlined system is `r format(total.post.low.years, big.mark=',')` years.
The number of years that could be saved is `r format(saved.low, big.mark=',')` years.

If we apply a full-time researchers assistant's time at an annual salary of \$`r format(salary, big.mark=',')`, then we have a salary cost of \$`r format(round(saved.low*salary/(10^6)), big.mark=',')` million per year.

# References

* Duplancic, C. , Crough, T. and Bell, S. C (2019) Multi‐centre ethics and research governance review can impede non‐interventional clinical research. _Intern Med J_, __49__: 722-728. doi:10.1111/imj.14158

* Robyn Clay-Williams, Natalie Taylor and Jeffrey Braithwaite (2019) Potential solutions to improve the governance of multicentre health services research. _Med J Aust_ __208__ (4). doi:10.5694/mja16.01268 

* National Health and Medical Research Council. Report on the Activity of Human Research Ethics Committees and Certified Institutions for the period: 1 January 2016 to 31 December 2016

* Adrian G. Barnett, Megan J. Campbell, Carla Shield, Alison Farrington, Lisa Hall, Katie Page, Anne Gardner, Brett G. Mitchell & Nicholas Graves (2016) The high costs of getting ethical and site-specific approvals for multi-centre research. _Research Integrity and Peer Review_ __1__ 16
